Deciding Bitvector Termination with Program Synthesis

### 2014-07-03 papers/termination/paper.tex
Simplify figure 4 slightly
revising kalashnikov
Correct big example
"Out of scope" -> "Standard"

\noindent {\bf Composing the environment with a loop.}
While proving that loop $L$ terminates, we create constraints on the environment in the
form of the predicate $I$ and the summary relation $S$.  For each loop $L'$ nested in the body
of $L$, we must add some more constraints saying that $S$ overapproximates $L'$.
Similarly, for each loop $L''$ appearing before $L$ in the program's control flow, we
must add constraints saying that $I$ overapproximates the states reachable via $L''$.
The details of how to create these constraints are beyond the scope of this paper, so
rather than presenting the algorithm for generating the constraints, we direct the
reader's attention to Figure~\ref{fig:environment-model}.  This figure contains a non-trivial program
along with the constraint system
encoding the termination of the outer loop.  Since the program contains sequential composition
of loops and nested loops, the constraint
system involves termination invariants $W_1, W_2, W_3$ and the inner loop summary $S$.
Note that despite its complexity, the constraint system is still a formula in the synthesis fragment.
Since the outer loop of this program does indeed terminate, the system is satisfiable, as witnessed by:
\begin{align*}
W_1(\langle v, x, y, z \rangle ) & \triangleq x \leq y + 1 \\
W_2(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
W_1(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge z' + v' = x \\
R(x) & = x
\end{align*}

\begin{figure*}
\begin{minipage}{0.2\textwidth}
\begin{lstlisting}
x = 0;
while (x <= y)
  x++;

@@

  x -= z;
  y -= z;
}
\end{lstlisting}
\end{minipage}
\vline
\begin{minipage}{0.8\textwidth}
\begin{align*}
 \exists W_1, W_2, W_3, R, S . \forall v, x, y, z, v', x', y', z' . \\
   x = 1 & \rightarrow W_1(v, x, y, z) \, \wedge \\
   x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
   x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x+1, y, z) \, \wedge \\
   x > 0 \wedge W_2(v, x, y, z) & \rightarrow W_3(0, x, y, x) \, \wedge \\
   v < y \wedge W_3(v, x, y, z) & \rightarrow W_3(v+1, x, y, z-1) \, \wedge \\
   v \geq y \wedge W_3(v, x, y, z) & \rightarrow W_2(v, x-z, y-z, z) \, \wedge \\
   S(\langle v, x, y, z, \rangle, \langle v', x', y', z' \rangle) \wedge v < y & \rightarrow S(\langle v, x, y, z \rangle, \langle v+1, x, y, z-1 \rangle) \, \wedge \\
   x > 0 \wedge W_2(v, x, y, z) & \rightarrow S(\langle v, x, y, z \rangle, \langle 0, x, y, x \rangle) \, \wedge \\
   x > 0 \wedge W_2(v, x, y, z) \wedge S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' \geq y' & \rightarrow \\
   R(v, x, y, z) > 0 \wedge R(v, x, y, z) > R(v', x' - z', y'-z', z')
\end{align*}
\end{minipage}
\caption{A non-trivial program and its constraint system\label{fig:environment-model}}
\end{figure*}

@@

\subsection{Genetic Programming and Incremental Evolution}

>>>>

\noindent {\bf Composing the environment with a loop.}
While proving that loop $L$ terminates, we create constraints on the environment in the
form of the predicate $I$ and the summary relation $S$.  For each loop $L'$ nested in the body
of $L$, we must add some more constraints saying that $S$ overapproximates $L'$.
Similarly, for each loop $L''$ appearing before $L$ in the program's control flow, we
must add constraints saying that $I$ overapproximates the states reachable via $L''$.
The details of how to create these constraints are standard, so
rather than presenting the algorithm for generating the constraints, we direct the
reader's attention to Figure~\ref{fig:environment-model}.  This figure contains a non-trivial program
along with the constraint system
encoding the termination of the outer loop.  Since the program contains sequential composition
of loops and nested loops, the constraint
system involves termination invariants $W_1, W_2, W_3$ and the inner loop summary $S$.
Note that despite its complexity, the constraint system is still a formula in the synthesis fragment.
Since the outer loop of this program does indeed terminate, the system is satisfiable, as witnessed by:
\begin{align*}
W_1(\langle v, x, y, z \rangle ) & \triangleq x \leq y + 1 \\
W_2(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge z' + v' = x' \wedge v' \leq y' \\
R(x) & = x
\end{align*}

\begin{figure*}
\begin{framed}
\begin{minipage}{0.19\textwidth}
\begin{lstlisting}
x = 0;
while (x <= y)
  x++;

@@

  x -= z;
  y -= z;
}
\end{lstlisting}
\end{minipage}
\vline
\begin{minipage}{0.82\textwidth}
\begin{align*}
 \exists W_1, W_2, R, S . \forall v, x, y, z, v', x', y', z' . \\
   x = 1 & \rightarrow W_1(v, x, y, z) \, \wedge \\
   x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
   x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x+1, y, z) \, \wedge \\
   %x > 0 \wedge W_2(v, x, y, z) & \rightarrow W_3(0, x, y, x) \, \wedge \\
   %v < y \wedge W_3(v, x, y, z) & \rightarrow W_3(v+1, x, y, z-1) \, \wedge \\
   %v \geq y \wedge W_3(v, x, y, z) & \rightarrow W_2(v, x-z, y-z, z) \, \wedge \\
   S(\langle v, x, y, z, \rangle, \langle v', x', y', z' \rangle) \wedge v < y & \rightarrow S(\langle v, x, y, z \rangle, \langle v+1, x, y, z-1 \rangle) \, \wedge \\
   x > 0 \wedge W_2(v, x, y, z) & \rightarrow S(\langle v, x, y, z \rangle, \langle 0, x, y, x \rangle) \, \wedge \\
   x > 0 \wedge W_2(v, x, y, z) \wedge S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' \geq y' & \rightarrow \\
   & W_2(v', x' - z', y' - z', z') \, \wedge \\
   & R(v, x, y, z) > 0 \, \wedge \\
   & R(v, x, y, z) > R(v', x' - z', y'-z', z')
\end{align*}
\end{minipage}
\caption{A non-trivial program and its constraint system\label{fig:environment-model}}
\end{framed}
\end{figure*}

@@

\subsection{Genetic Programming and Incremental Evolution} \label{sec:gp}
